---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# EFAtools

<!-- badges: start -->
[![Travis build status](https://travis-ci.com/mdsteiner/EFAtools.svg?branch=master)](https://travis-ci.com/mdsteiner/EFAtools)
<!-- badges: end -->

The EFAtools package provides functions to perform exploratory factor analysis (EFA) procedures and compare their solutions. The goal is to provide state-of-the-art factor retention methods and a high degree of flexibility in the EFA procedures. This way, implementations from R psych and SPSS can be compared. Moreover, functions for Schmid-Leiman transformation, and computation of omegas are provided. To speed up the analyses, some of the iterative procedures like principal axis factoring (PAF) are implemented in C++.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

```{r eval=FALSE}
install.packages("devtools")
devtools::install_github("mdsteiner/EFAtools")
```

To also build the vignette, use:

```{r eval=FALSE}
install.packages("devtools")
devtools::install_github("mdsteiner/EFAtools", build_vignettes = TRUE)
```

## Example

Here are a few examples on how to perform the analyses with the different types and how to compare the results using the `COMPARE` function. For more details, see the vignette by running `vignette("EFAtools", package = "EFAtools")`. The vignette provides a high-level introduction into the functionalities of the package.

```{r fig.height=3}
# load the package
library(EFAtools)

# Run all possible factor retention methods
N_FACTORS(test_models$baseline$cormat, N = 500, method = "ML")

# A type SPSS EFA to mimick the SPSS implementation with
# promax rotation
EFA_SPSS <- EFA(test_models$baseline$cormat, n_factors = 3, type = "SPSS",
                  rotation = "promax")

# look at solution
EFA_SPSS

# A type psych EFA to mimick the psych::fa() implementation with
# promax rotation
EFA_psych <- EFA(test_models$baseline$cormat, n_factors = 3, type = "psych",
                  rotation = "promax")

# compare the type psych and type SPSS implementations
COMPARE(EFA_SPSS$rot_loadings, EFA_psych$rot_loadings,
        x_labels = c("SPSS", "psych"))

# Perform a Schmid-Leiman transformation
SL <- SL(EFA_psych)

# Compute omegas from the Schmid-Leiman solution
OMEGA(SL, factor_corres = rep(c(3, 2, 1), each = 6))
```

## Contribute or Report Bugs

If you want to contribute or report bugs, please open an issue on GitHub or email us...
