---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# EFAtools

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/EFAtools)](https://CRAN.R-project.org/package=EFAtools)
[![R-CMD-check](https://github.com/mdsteiner/EFAtools/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mdsteiner/EFAtools/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/mdsteiner/EFAtools/graph/badge.svg)](https://app.codecov.io/gh/mdsteiner/EFAtools)
[![DOI](https://joss.theoj.org/papers/10.21105/joss.02521/status.svg)](https://doi.org/10.21105/joss.02521)

<!-- badges: end -->

The EFAtools package performs exploratory factor analysis (EFA) and compares EFA solutions. It offers current factor retention methods and many estimation, rotation, and correlation options. The package implements core iterative procedures, including principal axis factoring (PAF), rotation, and polychoric correlation estimation, in C++ to increase speed.

## Installation

You can install the release version from [CRAN](https://cran.r-project.org/) with:
```{r eval=FALSE}
install.packages("EFAtools")
```

You can install the development version from [GitHub](https://github.com/) with:

```{r eval=FALSE}
# install.packages("pak")
pak::pak("mdsteiner/EFAtools")
```

## Package Overview

The `efa_*` functions cover the steps of an EFA workflow:

* **Screening and suitability**: `efa_screen()` checks the data for multivariate normality, outliers, and suitability for factor analysis. Kaiser-Meyer-Olkin criterion and Bartlett's test of sphericity can also be run separately with `efa_kmo()` and `efa_bartlett()`.
* **Factor retention**: `efa_retain()` runs several factor retention criteria with a single call. They are also available on their own: `efa_cd()`, `efa_ekc()`, `efa_hull()`, `efa_kgc()`, `efa_map()`, `efa_nest()`, `efa_parallel()`, `efa_scree()`, and `efa_smt()`.
* **Fitting**: `efa_fit()` fits the factor model. `estimate_control()` and `rotate_control()` configure the estimation and rotation settings.
* **Special rotation and transformation**: `efa_procrustes()` rotates a solution onto a target. `efa_schmid_leiman()` transforms an oblique solution into a hierarchical one. 
* **Reliability**: `efa_reliability()` computes reliability and common-variance coefficients for a factor solution.
* **Factor scores**: `efa_scores()` estimates factor scores together with score-quality diagnostics.
* **Comparison**: `efa_compare()` compares two solutions (loadings or communalities).
* **Averaging**: `efa_average()` averages solutions across implementations and methods to assess their stability.
* **Multiple groups**: `efa_group()` fits a solution per group and compares them.
* **Multiple imputation**: `efa_mi()` fits and pools solutions across multiply imputed data sets.
* **Simulation**: `efa_simulate()` simulates data from a common-factor population model.
* **Power**: `efa_power()` performs analytic and simulation-based power analysis.

The uppercase names of the older package versions (`EFA()`, `N_FACTORS()`, ...) still work and keep their arguments. Use the `efa_*` names for new code.

The following vignettes and articles cover these in detail:

* [EFAtools](https://mdsteiner.github.io/EFAtools/articles/EFAtools.html): a complete EFA workflow, from screening to reliability and factor scores.
* [Migrating to the efa_* interface](https://mdsteiner.github.io/EFAtools/articles/Migrating_to_efa.html): what the old names and arguments map to.
* [EFA with ordinal and missing data](https://mdsteiner.github.io/EFAtools/articles/Ordinal_and_missing_data.html): polychoric and tetrachoric correlations, DWLS, FIML, and multiply imputed data.
* [Multigroup exploratory factor analysis](https://mdsteiner.github.io/EFAtools/articles/Multigroup_EFA.html): comparing solutions across groups and assessing approximate invariance.
* [Simulating data and power analysis for EFA](https://mdsteiner.github.io/EFAtools/articles/Simulation_and_power.html): simulating data and planning sample sizes.

## Examples

The following examples show some EFAtools functions.

```{r fig.height=3}
# load the package
library(EFAtools)
```

### Factor Retention

Use `efa_retain()` to test if your data are suitable for factor analysis. It runs multiple factor retention criteria in a single call.

With raw data, `efa_retain()` runs every criterion:

```{r fig.height=3}

# Run multiple factor retention methods
efa_retain(GRiPS_raw)
```

With a correlation matrix, `efa_retain()` skips the criteria that need raw data:

```{r}

efa_retain(DOSPERT$cormat, N = DOSPERT$N)

```

### EFA

#### Raw-Data Input

With raw data, you can use every feature: sandwich and bootstrap standard errors, DWLS estimation with polychoric correlations, and two-stage FIML estimation of correlations.

In the first example below, which uses bootstrap SEs (`se = "np-boot"`), the confidence intervals are percentile intervals over refitted resamples. For the loadings and factor correlations, the intervals are centred on the point estimate. For the indices derived from the chi-square (RMSEA, AIC, BIC, ECVI), the intervals are located above the point estimate, because each resample carries the sample's own misfit plus fresh sampling noise. So a point estimate can fall below its own lower bound in some cases. CFI and TLI are unaffected because they are ratios: their baseline chi-square shifts together with the model chi-square.

```{r fig.height=3}

# ULS / MINRES estimation with oblimin rotation and bootstrap SEs
mod <- efa_fit(DOSPERT_raw, n_factors = 5, estimator = "uls", rotation = "oblimin",
               se = "np-boot", seed = 1)
mod
# detailed output with summary()
summary(mod)

# inspect residuals with residuals()
residuals(mod)

# DWLS estimation based on polychoric correlations, with robust sandwich SEs
mod <- efa_fit(GRiPS_raw, n_factors = 1, estimator = "dwls", cor_method = "poly",
               se = "sandwich")
mod
summary(mod)
```


#### Correlation Input

When you do not have raw data, you can enter a correlation matrix and sample size instead. With ML estimation, you can still get information-based SEs (from the expected information matrix), but these assume multivariate normality.

```{r fig.height=3}

# ML estimation with oblimin rotation and information SEs, based on correlation
# matrix and N
mod <- efa_fit(test_models$baseline$cormat, N = 500,  n_factors = 3, estimator = "ml",
           rotation = "oblimin", se = "information")
mod
summary(mod)


```

## Citation

If you use this package in your research, please acknowledge it by citing: 

Steiner, M.D., & Grieder, S.G. (2020). EFAtools: An R package with fast and flexible implementations of exploratory factor analysis tools. *Journal of Open Source Software*, *5*(53), 2521. https://doi.org/10.21105/joss.02521

## Contribute or Report Bugs

If you want to contribute or report bugs, please open an issue on GitHub or email us at markus.d.steiner@gmail.com or silvia.steiner.grieder@gmail.com.
