---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# EFAtools

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/EFAtools)](https://CRAN.R-project.org/package=EFAtools)
[![R-CMD-check](https://github.com/mdsteiner/EFAtools/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mdsteiner/EFAtools/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/mdsteiner/EFAtools/graph/badge.svg)](https://app.codecov.io/gh/mdsteiner/EFAtools)
[![DOI](https://joss.theoj.org/papers/10.21105/joss.02521/status.svg)](https://doi.org/10.21105/joss.02521)

<!-- badges: end -->

The EFAtools package provides functions to perform exploratory factor analysis (EFA) procedures and compare their solutions. The goal is to provide state-of-the-art factor retention methods and a high degree of flexibility in the EFA procedures. To speed up the analyses, some of the iterative procedures like principal axis factoring (PAF) are implemented in C++.

## Installation

You can install the release version from [CRAN](https://cran.r-project.org/) with:
```{r eval=FALSE}
install.packages("EFAtools")
```

You can install the development version from [GitHub](https://github.com/) with:

```{r eval=FALSE}
# install.packages("pak")
pak::pak("mdsteiner/EFAtools")
```


## Examples

Here are a few examples of EFAtools functionalities

```{r fig.height=3}
# load the package
library(EFAtools)
```

### Factor Retention

Use `N_FACTORS()` to test the suitability of the data for factor analysis and to run multiple factor retention criteria with a single function call.

Works on raw-data:

```{r fig.height=3}

# Run multiple factor retention methods
N_FACTORS(GRiPS_raw)
```

Skips some criteria on correlation input:

```{r}

N_FACTORS(DOSPERT$cormat, N = DOSPERT$N)

```

### EFA

#### Raw-Data Input

With raw-data input, you can use all functionalities, including sandwich and bootstrap standard errors and DWLS estimation with polychoric correlations or two-stage FIML estimation of correlations.


```{r fig.height=3}

# ULS / MINRES estimation with oblimin rotation and sandwich SEs
mod <- EFA(DOSPERT_raw, n_factors = 5, method = "ULS", rotation = "oblimin",
           se = "sandwich")
mod
# detailed output with summary()
summary(mod)

#inspect residuals with residuals()
residuals(mod)

# DWLS estimation based on polychoric correlations, with robust sandwich SEs
mod <- EFA(GRiPS_raw, n_factors = 1, method = "DWLS", cor_method = "poly",
           se = "sandwich")
mod
summary(mod)
```


#### Correlation Input

When you don't have raw data available, you can enter a correlation matrix and sample size. With ML-estimation, you can still get information SEs, but note that they assume multivariate normality.

```{r fig.height=3}

# ML estimation with oblimin rotation and information SEs, based on correlation
# matrix and N
mod <- EFA(test_models$baseline$cormat, N = 500,  n_factors = 3, method = "ML",
           rotation = "oblimin", se = "information")
mod
summary(mod)


```

## Citation

If you use this package in your research, please acknowledge it by citing: 

Steiner, M.D., & Grieder, S.G. (2020). EFAtools: An R package with fast and flexible implementations of exploratory factor analysis tools. *Journal of Open Source Software*, *5*(53), 2521. https://doi.org/10.21105/joss.02521

## Contribute or Report Bugs

If you want to contribute or report bugs, please open an issue on GitHub or email us at markus.d.steiner@gmail.com or silvia.steiner.grieder@gmail.com.
