---
title: "EFAtools"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EFAtools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 7,
  fig.align = "center"
)
```

`EFAtools` provides a complete workflow for exploratory factor analysis (EFA). Starting
from raw data or a correlation matrix, it takes you through every stage of a typical
analysis:

1. **Screening** the data for factorability, multicollinearity, non-normality, and
   outliers (`efa_screen()`, with `efa_bartlett()` and `efa_kmo()` for the individual
   measures).
2. **Retention**: deciding how many factors to keep, with a comprehensive suite of factor
   retention criteria (`efa_retain()` and the individual criteria it wraps).
3. **Extraction and rotation** of the factor solution (`efa_fit()`), with a choice
   of estimators and orthogonal or oblique rotations.
4. **Post-processing**: comparing solutions (`efa_compare()`), averaging across analytic
   choices (`efa_average()`), estimating factor scores (`efa_scores()`), a Schmid-Leiman
   transformation (`efa_schmid_leiman()`), and reliability coefficients including
   McDonald's omegas (`efa_reliability()`).

The computationally intensive routines are implemented in C++ for speed. This vignette
walks through the whole workflow on a single data set; it is an overview rather than a
tutorial on the methods themselves, so please consult the individual help pages for the
statistical details and literature references.

The package can be installed from CRAN using `install.packages("EFAtools")`, or from GitHub
using `pak::pak("mdsteiner/EFAtools")`, and then loaded using:

```{r}
library(EFAtools)
```

We use the `DOSPERT_raw` data set throughout, which contains responses of 3123 participants
to the Domain Specific Risk Taking Scale (DOSPERT); see `?DOSPERT_raw` for details. The
data are rather large, so, purely to keep this vignette quick to build, we work with the
first 500 observations. In a real analysis you would of course use the full data set.

```{r}
# only use a subset to make analyses faster
DOSPERT_sub <- DOSPERT_raw[1:500, ]
```

## Screening the Data

Before extracting factors it is worth checking whether the data are suitable for factor
analysis, and whether they satisfy the assumptions of the estimator you plan to use.
`efa_screen()` collects the relevant diagnostics in one place. From raw data it reports the
Kaiser-Meyer-Olkin (KMO) measure of sampling adequacy (overall and per variable), Bartlett's
test of sphericity, the determinant and condition number of the correlation matrix, the
squared multiple correlation (SMC) and per-variable diagnostics for each item, tests of
multivariate normality (Mardia and Henze-Zirkler), and multivariate outliers, and it closes
with a set of recommendations.

```{r, warning = FALSE}
efa_screen(DOSPERT_sub, seed = 2)
```

The overall KMO is meritorious and Bartlett's test is clearly significant, so the data are
factorable. The near-zero determinant flags some multicollinearity, which is expected for a
scale with strongly related items within each risk domain. More interesting are the last few
sections: three items have a sparse response category, and both the Mardia and the
Henze-Zirkler tests reject multivariate normality. A consequence is that normal-theory
standard errors and fit indices may be biased, so robust (sandwich) or bootstrapped standard
errors are preferable, and Bartlett's test should be read with caution. For a fully ordinal
treatment you would move to polychoric correlations with a categorical least-squares
estimator.

On the outlier section: The robust distances come from a minimum
covariance determinant (MCD) estimate, whose random subset search is the reason
`efa_screen()` takes a `seed`. The MCD assumes roughly elliptical data, and on markedly
non-normal data like these, large robust distances are common. This is why almost half of
the observations end up flagged (On such heavily discretised data the most robust 50%
subset can also turn out singular, in which case `efa_screen()` falls back to classical
Mahalanobis distances and labels the output accordingly). The flagged cases are therefore
better understood as evidence of that non-normality than as genuine outliers to remove.
Inspect them (via `$outliers$flagged`) rather than deleting them automatically.

If you only want the two classic factorability tests, they are also available on their own.
Bartlett's test of sphericity checks whether the correlation matrix differs from an identity
matrix (it should be significant), and the KMO criterion summarises how strongly each variable
is predicted by the others:

```{r}
# Bartlett's test of sphericity
efa_bartlett(DOSPERT_sub)

# KMO criterion
efa_kmo(DOSPERT_sub)
```

Both tests can also be requested directly inside `efa_retain()`, alongside the factor
retention criteria discussed next.

## Deciding How Many Factors to Retain

One of the most consequential decisions in an EFA is how many factors to extract. There is a
large number of factor retention criteria, and no single one dominates the others: which
criterion works best depends on the structure of the data — how many indicators there are,
how strong the factors are, and how strongly they intercorrelate. The current recommendation
is therefore to examine several criteria together and check whether they converge (see, for
example, [Auerswald and Moshagen, 2019](https://doi.org/10.1037/met0000200)).

`EFAtools` implements a broad set of these criteria. They can be called individually, or all
(or a selection) of them together through `efa_retain()`.

### Calling Individual Criteria

Each criterion has its own function. For a parallel analysis based on squared multiple
correlations (SMC; also called parallel analysis with principal factors), for example, you
would run:

```{r}
# parallel analysis based on SMC eigenvalues
pa <- efa_parallel(DOSPERT_sub, eigen_type = "smc")
pa
```

Printing the result reports the suggested number of factors. Each criterion also has a
`plot()` method that draws the figure behind the decision — for a parallel analysis, the
observed eigenvalues against the reference eigenvalues from the simulated data:

```{r}
plot(pa)
```

Other criteria work the same way. The empirical Kaiser criterion, for instance:

```{r}
# empirical Kaiser criterion
efa_ekc(DOSPERT_sub)
```

The following criteria are currently implemented: comparison data (`efa_cd()`), the empirical
Kaiser criterion (`efa_ekc()`), the hull method (`efa_hull()`), the Kaiser-Guttman criterion
(`efa_kgc()`), the minimum average partial test (`efa_map()`), the next eigenvalue sufficiency
test (`efa_nest()`), parallel analysis (`efa_parallel()`), the scree test (`efa_scree()`), and
sequential model tests (`efa_smt()`). Many of them offer several variants of the underlying
method. For instance, parallel analysis, can be based on eigenvalues from unity (principal
components), SMCs, or a full EFA. See the individual help pages for the options.

### Running Several Criteria at Once with `efa_retain()`

To compare criteria, it is easier to call `efa_retain()`, a wrapper around the implemented
retention criteria that also runs Bartlett's test and computes the KMO. You can pick a subset
with the `criteria` argument:

```{r}
ret <- efa_retain(DOSPERT_sub,
                  criteria = c("parallel", "ekc", "kgc", "smt", "map"))
ret
```

The printed summary is text only, so the scree test's suggestion to inspect the plot needs an
explicit call. `plot()` on the result draws the figure for every criterion that has one:

```{r}
plot(ret)
```

Omitting `criteria` runs the full suite, including the hull method, the next eigenvalue
sufficiency test, and comparison data; the latter is computationally expensive and can be
slow on larger data sets, which is why we leave it out here.

This is the situation one would rather avoid, but it does happen: the criteria disagree, with
suggestions ranging from four to well above ten factors. When there is no clear convergence,
the choice becomes partly a matter of judgement. We proceed with six factors, consistent with the minimum average partial test and the RMSEA-based sequential model test above (note that this DOSPERT version assumes 5 factors, 6 factors splits the financial risk items into two factors).

All criteria except comparison data (which needs raw data) can also be applied to a
correlation matrix, in which case the sample size must be supplied. On a cleaner data set the
criteria often agree:

```{r}
efa_retain(test_models$baseline$cormat, N = 500, estimator = "uls",
           criteria = c("parallel", "ekc", "smt"),
           eigen_type_other = c("smc", "pca"))
```

Here the criteria agree on three factors, with only the conservative RMSEA lower-bound
variant of the sequential model test suggesting two.

## Extracting Factors with `efa_fit()`

`efa_fit()` performs the factor extraction and rotation. For extraction you can choose
principal axis factoring (PAF), maximum likelihood (ML), or unweighted least squares (ULS,
also known as MINRES); diagonally weighted least squares (DWLS) is additionally available
for ordinal data analysed with polychoric or tetrachoric correlations (see the
[EFA with ordinal and missing data](Ordinal_and_missing_data.html) vignette). For rotation
the package offers varimax and promax as well as a range
of further orthogonal and oblique rotations (e.g., quartimax, equamax, oblimin, quartimin,
geomin, bentler, bifactor, and simplimax), all computed by rotation engines built into the
package.

An EFA with PAF and no rotation is as simple as:

```{r}
efa_fit(DOSPERT_sub, n_factors = 6)
```

To rotate the loadings, set the `rotation` argument. Here we use a promax rotation. We assign the
result, as later sections reuse this solution:

```{r}
efa_dospert <- efa_fit(DOSPERT_sub, n_factors = 6, rotation = "promax")
efa_dospert
```

We can also use a different estimator and rotation, here for example ULS and an oblimin rotation:

```{r}
efa_uls <- efa_fit(DOSPERT_sub, n_factors = 6, rotation = "oblimin", estimator = "uls")
efa_uls
```

Note that `estimator = "uls"` was written in lowercase although the canonical spelling is
`"ULS"`. This holds throughout the `efa_*` interface: wherever an argument takes a fixed set
of values, capitalization does not matter, while the result stores and prints the canonical
spelling. For an argument that takes a single value an unambiguous abbreviation is enough as
well, and a value matching nothing raises an error naming the argument and listing the
choices. Arguments that take several values at once, such as the analytic choices
`efa_average()` averages over, need the full spelling of each value.

## Comparing Solutions with `efa_compare()`

`efa_compare()` provides a quick way to see how similar two loading (pattern) matrices are —
whether they come from different implementations, different estimators, or different rotations.
For instance, to compare the PAF/promax solution with the ULS/oblimin one:

```{r}
efa_compare(efa_dospert$rot_loadings, efa_uls$rot_loadings,
            x_labels = c("PAF and promax", "ULS and oblimin"))
```

## Averaging Across Analytic Choices with `efa_average()`

Comparing two solutions by hand answers the question for those two. `efa_average()` asks it
across a whole grid at once: it runs an EFA for every combination of the settings you allow it
to vary, averages the loadings across the solutions that converged, and reports the range each
loading spans. That range is the useful part — it shows how much of the result is the data and
how much is the analytic choice. Here we vary only the estimator, keeping one implementation
each of PAF, ULS, and ML with a promax rotation:

```{r}
avg_dospert <- efa_average(DOSPERT_sub, n_factors = 6,
                           estimator = c("paf", "uls", "ml"),
                           type = "EFAtools", start_method = "psych")
avg_dospert
```

Left unrestricted, the defaults sweep the different implementations of principal axis
factoring and promax rotation instead, which is considerably more work; see `?efa_average` for
the settings that can be varied and for `plot()`, which shows the minimum, maximum, and
average loading of every indicator.

## Factor Scores with `efa_scores()`

If you need factor scores, `efa_scores()` estimates them directly from an `efa_fit()` solution, and reports score-quality diagnostics (determinacy,
univocality, and the Guttman indeterminacy index):

```{r}
fac_scores <- efa_scores(DOSPERT_sub, f = efa_dospert)
fac_scores
```

## Schmid-Leiman Transformation and McDonald's Omegas

For the Schmid-Leiman transformation and the omega coefficients we return to the PAF/promax
solution (`efa_dospert`) from above. Its indicator names encode the risk domain each item
belongs to (ethical, financial, health, recreational, and social), and its pattern
coefficients recovered these theoretical domains fairly well: items from the same domain
tend to load on the same factor. The factor intercorrelations also included some sizeable
values, which raises the question of whether a general factor underlies the domains. A
Schmid-Leiman transformation lets us examine that.

### Schmid-Leiman Transformation

The Schmid-Leiman (SL) transformation orthogonalises an oblique solution into a hierarchical
one with a general factor and orthogonalised group factors. `efa_schmid_leiman()` performs it:

```{r}
sl_dospert <- efa_schmid_leiman(efa_dospert)
sl_dospert
```

The general-factor column shows that the items load substantially on the general factor
across all domains except the social one, whose items load least strongly on it on average
(mean general-factor loading around .19, against roughly .38 to .55 in the other domains) —
that is, the non-social domains covary considerably, while the social domain stands somewhat
apart. Individual social items vary considerably around that average, so this is a statement
about the domain and not about every item in it.

### McDonald's Omegas

Finally, `efa_reliability()` computes omega estimates and related indices from the SL solution.
You can either specify the indicator-to-factor correspondences via the `factor_map` argument or
let the function determine them automatically, in which case each indicator is assigned to its
highest-loading factor. With a reasonably simple structure here, it is easiest to leave
`factor_map` unspecified — a few items do land on a neighbouring domain's factor, which is
exactly what motivates the explicit map shown afterwards:

```{r}
efa_reliability(sl_dospert)
```

To set the correspondences explicitly instead — for example to match a theoretical assignment —
pass a `factor_map` matrix:

```{r}
efa_reliability(sl_dospert,
  factor_map = matrix(c(rep(1, 6), rep(0, 24),                  # F1: ethical
                        rep(0, 18), rep(1, 6), rep(0, 6),       # F2: recreational
                        rep(0, 6), 1, 0, 1, 0, 1, rep(0, 19),   # F3: financial (1, 3, 5)
                        rep(0, 12), rep(1, 6), rep(0, 12),      # F4: health
                        rep(0, 24), rep(1, 6),                  # F5: social
                        rep(0, 7), 1, 0, 1, 0, 1, rep(0, 18)),  # F6: financial (2, 4, 6)
                      ncol = 6, byrow = FALSE))
```

The columns of `factor_map` are matched to the
group factors **by position**, so the assignment has to be read off the solution rather than
assumed: a column naming the wrong set of items yields a well-formed but meaningless subscale
omega instead of an error. Comparing the two tables, the health row is the one that moves most:
three of the six health items load higher on the ethical factor than on the health one, so
pinning all six to F4 lowers that subscale's omega relative to the automatic assignment.

## Where to Next

This vignette covered the core workflow. `EFAtools` goes further in several directions, each
of which has its own vignette or article:

* [EFA with ordinal and missing data](Ordinal_and_missing_data.html): analysing ordinal
  items with polychoric correlations and handling missing data.
* [Migrating to the efa_* interface](Migrating_to_efa.html): a guide for users coming from
  the older uppercase function names.
* [Multigroup EFA](https://mdsteiner.github.io/EFAtools/articles/Multigroup_EFA.html)
  (on the package website): comparing factor structures across groups.
* [Simulation and power](https://mdsteiner.github.io/EFAtools/articles/Simulation_and_power.html)
  (on the package website): simulating data and running power analyses for EFA.

Run `browseVignettes("EFAtools")` to see the vignettes installed with the package, or visit
the [package website](https://mdsteiner.github.io/EFAtools/) for the full set of articles.
