---
title: "EFA with ordinal and missing data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EFA with ordinal and missing data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 7,
  fig.align = "center"
)

# The multiple-imputation section uses mice to create the imputations. It is only
# suggested by the package, so the imputation chunks are evaluated only when it is
# installed.
mice_ok <- requireNamespace("mice", quietly = TRUE)
```

Two features of real data complicate an exploratory factor analysis: items are often
**ordinal** (a handful of Likert categories rather than a continuous scale), and some
responses are usually **missing**. `EFAtools` handles both within the ordinary `efa_fit()`
workflow, without switching packages. This vignette shows how. It assumes familiarity with
the basic workflow covered in the [EFAtools](EFAtools.html) vignette and focuses on what
changes for ordinal and incomplete data.

```{r}
library(EFAtools)
```

So that the examples are self-contained and reproducible, we generate the data with
`efa_simulate()` from a known three-factor population (18 indicators, six per factor, with
moderately correlated factors), using fixed seeds throughout.

```{r}
Lambda <- population_models$loadings$baseline  # 18 x 3 loading pattern
Phi    <- population_models$phis_3$moderate    # moderate factor intercorrelations
```

## Ordinal Data

Rating-scale items are not continuous: they take a few ordered values, and a Pearson
correlation between two such items underestimates the association between the underlying
constructs. The polychoric correlation instead estimates the correlation of the continuous
latent variables assumed to underlie the observed categories, and pairing it with a
categorical estimator removes the bias that treating the items as continuous introduces.

We draw 400 responses on a four-category scale. Because the latent data are normal, cutting
them at the standard-normal category thresholds already leaves the population *polychoric*
correlation of the discretised data equal to the target correlation; `match = "polychoric"`
records that this is what we are after, and would reject the request had we asked for
non-normal marginals.

```{r}
d_ord <- efa_simulate(N = 400, Lambda = Lambda, Phi = Phi,
                      categories = 4, match = "polychoric", seed = 2024)$data
d_ord[1:5, 1:6]
```

### Screening Ordinal Data

`efa_screen()` reports, among its diagnostics, how many response categories each item has
and whether the data are multivariate normal — the two things that decide whether an
ordinal treatment is worthwhile.

```{r, warning = FALSE}
efa_screen(d_ord, seed = 42)
```

The sampling adequacy (KMO) and sphericity checks confirm the data are factorable. The
telling parts are the multivariate-normality section and the recommendations: Mardia's
kurtosis and the Henze-Zirkler test reject normality, and the recommendations flag that every
item has fewer than five response categories. Together these spell out the consequence — with
few categories and non-normal data, a polychoric correlation with a categorical estimator
such as DWLS is less biased than normal-theory maximum likelihood, and the normal-theory
standard errors and fit indices are better replaced by robust (sandwich) versions.

### Polychoric Correlations, DWLS, and Robust Standard Errors

`efa_fit()` computes the polychoric correlation when `cor_method = "poly"` and fits it with
diagonally weighted least squares when `estimator = "DWLS"` — the estimator recommended for
ordinal data, which weights each correlation residual by the inverse asymptotic variance of
the corresponding polychoric correlation. Requesting `se = "sandwich"` adds robust standard
errors and a scaled (Satorra-Bentler) chi-square that stay valid under the non-normality
these data show.

```{r}
efa_poly <- efa_fit(d_ord, n_factors = 3, cor_method = "poly", estimator = "dwls",
                    rotation = "oblimin", se = "sandwich")
efa_poly
```

The pattern matrix recovers the three factors cleanly (six indicators each), and the model
fit reports a **scaled** chi-square with its CFI, TLI, and RMSEA. Because the chi-square is
a scaled statistic, the AIC and BIC (which are defined on the unscaled likelihood
discrepancy) are left `NA`.

For binary items, `cor_method = "tetra"` computes tetrachoric correlations and runs the same
DWLS and sandwich machinery.

The robust standard errors accompany each estimated quantity; for the rotated loadings, for
example:

```{r}
round(efa_poly$SE$rot_loadings, 3)
```

The matching confidence intervals live in `efa_poly$CI`, and `summary(efa_poly)` prints them
as a labelled table alongside the model diagnostics.

### Why Not Just Treat the Items as Continuous?

To see what the ordinal treatment buys, fit the same data as if they were continuous — a
Pearson correlation with maximum likelihood — and compare the rotated loadings with
`efa_compare()`.

```{r}
efa_cont <- efa_fit(d_ord, n_factors = 3, cor_method = "pearson", estimator = "ML",
                    rotation = "oblimin")

cmp <- efa_compare(efa_poly$rot_loadings, efa_cont$rot_loadings,
                   x_labels = c("Polychoric / DWLS", "Pearson / ML"))
cmp
plot(cmp)
```

The two solutions agree on the structure, but the polychoric loadings are systematically a
little larger: treating the items as continuous attenuates the loadings, because the Pearson
correlation understates the latent associations. With only four categories here the gap is
modest, but it widens as the number of categories drops (it is largest for binary items) and
as the category thresholds grow more asymmetric (skewed items). This is why a polychoric or
tetrachoric treatment is preferable for genuinely ordinal items with few categories.

The polychoric route does make its own demands, though: it assumes a normal latent variable
underlies each item, and it needs an adequate sample size and reasonably populated
response-category combinations. When categories are very sparse (rare responses, small
samples), the polychoric asymptotic covariance behind the DWLS weights and the robust
standard errors becomes unreliable — `efa_fit()` warns when empty category combinations are
present — and collapsing rare categories can help.

## Missing Data

When some responses are missing, dropping every incomplete case (listwise deletion) wastes
data and can bias the results unless the values are missing completely at random. `EFAtools`
offers two principled alternatives that assume only that the data are missing at random
(MAR): a single-analysis route via full-information maximum likelihood, and a
multiple-imputation route via `efa_mi()`.

We simulate 250 continuous cases with about 15% of values missing at random, where each
item's missingness depends on another item's value. By default `efa_simulate()` holes every
column, so each item's MAR predictor would itself be partly missing: the mechanism is then
MAR given the *complete* data, but not ignorable for an analyst who sees only the observed
data, and estimators that are consistent under ignorable MAR keep a residual bias — a
property of the generator rather than of the estimators, growing with `missing_prop` and
`missing_strength`.

`missing_vars` avoids that. Here the first nine items carry the missing values and each is
driven by one of the last nine, which stay complete. Every predictor is therefore fully
observed, which is ignorably MAR — exactly the assumption the two routes below rely on.

```{r}
d_miss <- efa_simulate(N = 250, Lambda = Lambda, Phi = Phi,
                       missing = "MAR", missing_prop = 0.15,
                       missing_vars = 1:9, missing_predictor = 10:18,
                       seed = 2024)$data
round(mean(is.na(d_miss)), 3)          # overall proportion missing
round(colMeans(is.na(d_miss)), 3)      # holed items only
```

### Two-Stage Full-Information Maximum Likelihood

With `cor_method = "fiml"`, `efa_fit()` estimates the saturated mean and covariance from all
the observed data by an EM algorithm (assuming the data are MAR) and analyses the resulting
correlation — a single fit that uses every case rather than only the complete ones. The
model fit is reported as corrected two-stage (Satorra-Bentler) statistics.

```{r}
efa_fiml <- efa_fit(d_miss, n_factors = 3, cor_method = "fiml", estimator = "ml",
                    rotation = "oblimin")
efa_fiml
```

The solution again recovers the three factors, and the printout records that the correlation
was obtained by two-stage FIML. Standard errors are available here too: for `estimator = "ML"`
or `"ULS"`, `se = "information"` or `"sandwich"` return the corrected two-stage standard
errors, and `se = "np-boot"` works with any estimator.

### Multiple Imputation with `efa_mi()`

The alternative is to impute the missing values several times, fit each completed dataset,
and pool the results. `EFAtools` does not impute the data itself — use a dedicated tool such
as the [mice](https://CRAN.R-project.org/package=mice) package — but `efa_mi()` takes the
list of completed datasets and does the factor-analytic pooling.

Here we create five imputations with mice (a Bayesian linear-regression model, appropriate
for these continuous items) and collect them into a list.

```{r, eval = mice_ok}
imp <- mice::mice(as.data.frame(d_miss), m = 5, method = "norm",
                  printFlag = FALSE, seed = 123)
dat_list <- lapply(seq_len(imp$m), function(i) mice::complete(imp, i))
```

`efa_mi()` fits the same `efa_fit()` model to each imputed dataset — the extraction,
rotation, and standard-error options are passed through `...` — aligns the solutions to a
common factor space (rotation is only identified up to reflection and permutation, so the
imputations must be matched before averaging), and pools them.

```{r, eval = mice_ok}
efa_pooled <- efa_mi(dat_list, n_factors = 3, estimator = "ml", rotation = "oblimin")
efa_pooled
```

The pooled loadings recover the three factors. Point estimates are averaged across the
imputations after alignment. The model chi-square, and the RMSEA and AIC/BIC derived from
it, are pooled with the D2 rule — which is why the printout labels the pooled chi-square as a
D2 statistic with its own reference distribution — whereas the incremental CFI and TLI are
averaged across the per-imputation fits. Requesting standard errors in the call (for
example `se = "information"` or `se = "np-boot"`) additionally pools them with Rubin's rules,
so the between-imputation variability inflates the pooled standard errors. Because multiple
imputation propagates the extra uncertainty from the missing data, its pooled fit statistics
are not directly comparable with the single FIML fit above; read them together with the
per-imputation fits stored in the returned object.

Multiple imputation is not limited to continuous data: since `efa_mi()` forwards its
arguments to `efa_fit()`, imputed ordinal datasets can be pooled with `cor_method = "poly"`
and `estimator = "DWLS"` in exactly the same way.

Which route to prefer is largely practical. FIML is a single, efficient fit and is the
simpler default when the analysis model is the whole story. Multiple imputation is more
flexible when the imputation model should draw on auxiliary variables not in the factor
model, or when the same imputations feed several downstream analyses.

## Where to Next

This vignette covered the ordinal and missing-data extensions of the workflow. For the core
analysis — screening, factor retention, extraction and rotation, and the post-processing
tools — see the [EFAtools](EFAtools.html) vignette, and the individual help pages for the
statistical details and references. Run `browseVignettes("EFAtools")` for the vignettes
installed with the package, or visit the
[package website](https://mdsteiner.github.io/EFAtools/) for the full set of articles.
