---
title: "EFA with ordinal and missing data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EFA with ordinal and missing data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.align = "center"
)

# The multiple-imputation section uses mice to create the imputations. It is only
# suggested by the package, so the imputation chunks are evaluated only when it is
# installed.
mice_ok <- requireNamespace("mice", quietly = TRUE)
```

Two features of real data complicate an exploratory factor analysis: items are often
**ordinal** (a handful of Likert categories rather than a continuous scale), and some
responses are usually **missing**. `EFAtools` handles both within the ordinary `efa_fit()`
workflow, without switching packages. This vignette shows how. It assumes familiarity with
the basic workflow covered in the [EFAtools](EFAtools.html) vignette and focuses on what
changes for ordinal and incomplete data.

```{r}
library(EFAtools)
```

So that the examples are self-contained and reproducible, we generate the data with
`efa_simulate()` from a known three-factor population (18 indicators, six per factor, with
moderately correlated factors), using fixed seeds throughout.

```{r}
Lambda <- population_models$loadings$baseline  # 18 x 3 loading pattern
Phi    <- population_models$phis_3$moderate    # moderate factor intercorrelations
```

## Ordinal Data

Rating-scale items are not continuous: they take a few ordered values, and a Pearson
correlation between two such items underestimates the association between the underlying
constructs. The polychoric correlation instead estimates the correlation of the continuous
latent variables assumed to underlie the observed categories, and pairing it with a
categorical estimator removes the bias that treating the items as continuous introduces.

We draw 400 responses on a four-category scale. Because the latent data are normal, cutting
them at the standard-normal category thresholds already leaves the population *polychoric*
correlation of the discretised data equal to the target correlation; `match = "polychoric"`
records that this is what we are after, and would reject the request had we asked for
non-normal marginals.

```{r}
d_ord <- efa_simulate(N = 400, Lambda = Lambda, Phi = Phi,
                      categories = 4, match = "polychoric", seed = 2024)$data
d_ord[1:5, 1:6]
```

### Screening Ordinal Data

`efa_screen()` reports, among its diagnostics, how many response categories each item has
and whether the data are multivariate normal — the two things that decide whether an
ordinal treatment is worthwhile.

```{r, warning = FALSE}
efa_screen(d_ord, seed = 42)
```

The sampling adequacy (KMO) and sphericity checks confirm the data are factorable. The
telling parts are the multivariate-normality section and the recommendations: Mardia's
kurtosis and the Henze-Zirkler test reject normality, and the recommendations flag that every
item has fewer than five response categories. Together these spell out the consequence — with
few categories and non-normal data, a polychoric correlation with a categorical estimator
such as DWLS is less biased than normal-theory maximum likelihood, and the normal-theory
standard errors and fit indices are better replaced by robust (sandwich) versions.

### Polychoric Correlations, DWLS, and Robust Standard Errors

`efa_fit()` computes the polychoric correlation when `cor_method = "poly"` and fits it with
diagonally weighted least squares when `estimator = "DWLS"` — the estimator recommended for
ordinal data, which weights each correlation residual by the inverse asymptotic variance of
the corresponding polychoric correlation. Requesting `se = "sandwich"` adds robust standard
errors and a scaled (Satorra-Bentler) chi-square that stay valid under the non-normality
these data show.

```{r}
efa_poly <- efa_fit(d_ord, n_factors = 3, cor_method = "poly", estimator = "dwls",
                    rotation = "oblimin", se = "sandwich")
efa_poly
```

The pattern matrix recovers the three factors cleanly (six indicators each), and the model
fit reports a **scaled** chi-square with its CFI, TLI, and RMSEA. Because the chi-square is
a scaled statistic, the AIC and BIC (which are defined on the unscaled likelihood
discrepancy) are left `NA`.

For binary items, `cor_method = "tetra"` computes tetrachoric correlations and runs the same
DWLS and sandwich machinery.

The robust standard errors accompany each estimated quantity; for the rotated loadings, for
example:

```{r}
round(efa_poly$SE$rot_loadings, 3)
```

The matching confidence intervals live in `efa_poly$CI`, and `summary(efa_poly)` prints them
as a labelled table alongside the model diagnostics.

### Why Not Just Treat the Items as Continuous?

To see what the ordinal treatment buys, fit the same data as if they were continuous — a
Pearson correlation with maximum likelihood — and compare the rotated loadings with
`efa_compare()`.

```{r}
efa_cont <- efa_fit(d_ord, n_factors = 3, cor_method = "pearson", estimator = "ML",
                    rotation = "oblimin")

cmp <- efa_compare(efa_poly$rot_loadings, efa_cont$rot_loadings,
                   x_labels = c("Polychoric / DWLS", "Pearson / ML"))
cmp
plot(cmp)
```

The two solutions agree on the structure, but the polychoric loadings are systematically a
little larger: treating the items as continuous attenuates the loadings, because the Pearson
correlation understates the latent associations. With only four categories here the gap is
modest, but it widens as the number of categories drops (it is largest for binary items) and
as the category thresholds grow more asymmetric (skewed items). This is why a polychoric or
tetrachoric treatment is preferable for genuinely ordinal items with few categories.

The polychoric route does make its own demands, though: it assumes a normal latent variable
underlies each item, and it needs an adequate sample size and reasonably populated
response-category combinations. When categories are very sparse (rare responses, small
samples), the polychoric asymptotic covariance behind the DWLS weights and the robust
standard errors becomes unreliable — `efa_fit()` warns when empty category combinations are
present — and collapsing rare categories can help.

## Missing Data

When some responses are missing, dropping every incomplete case (listwise deletion) wastes
data and can bias the results unless the values are missing completely at random. `EFAtools`
offers two principled alternatives that assume only that the data are missing at random
(MAR): a single-analysis route via full-information maximum likelihood, and a
multiple-imputation route via `efa_mi()`.

We simulate 250 continuous cases with about 15% of values missing at random, where each
item's missingness depends on another item's value. By default `efa_simulate()` holes every
column, so each item's MAR predictor would itself be partly missing: the mechanism is then
MAR given the *complete* data, but not ignorable for an analyst who sees only the observed
data, and estimators that are consistent under ignorable MAR keep a residual bias — a
property of the generator rather than of the estimators, growing with `missing_prop` and
`missing_strength`.

`missing_vars` avoids that. Here the first nine items carry the missing values and each is
driven by one of the last nine, which stay complete. Every predictor is therefore fully
observed, which is ignorably MAR — exactly the assumption the two routes below rely on.

```{r}
d_miss <- efa_simulate(N = 250, Lambda = Lambda, Phi = Phi,
                       missing = "MAR", missing_prop = 0.15,
                       missing_vars = 1:9, missing_predictor = 10:18,
                       seed = 2024)$data
round(mean(is.na(d_miss)), 3)          # overall proportion missing
round(colMeans(is.na(d_miss)), 3)      # holed items only
```

### Two-Stage Full-Information Maximum Likelihood

With `cor_method = "fiml"`, `efa_fit()` estimates the saturated mean and covariance from all
the observed data by an EM algorithm (assuming the data are MAR) and analyses the resulting
correlation — a single fit that uses every case rather than only the complete ones. The
model fit is reported as corrected two-stage (Satorra-Bentler) statistics.

```{r}
efa_fiml <- efa_fit(d_miss, n_factors = 3, cor_method = "fiml", estimator = "ml",
                    rotation = "oblimin")
efa_fiml
```

The solution again recovers the three factors, and the printout records that the correlation
was obtained by two-stage FIML. Standard errors are available here too: for `estimator = "ML"`
or `"ULS"`, `se = "information"` or `"sandwich"` return the corrected two-stage standard
errors, and `se = "np-boot"` works with any estimator.

The correction itself needs a well-behaved saturated covariance. When that cannot be formed —
typically in a small sample with a high proportion of missing values, or with near-collinear
items — `efa_fit()` warns and keeps the plain two-stage likelihood-ratio statistic instead of
discarding the test. Such a fit is never presented as corrected: the printed chi-square line
is labelled **uncorrected**, the object carries
`fit_indices$chi_scaled_type == "uncorrected.lrt"`, and the accompanying *p*-value, CFI, TLI,
and RMSEA should be read as indicative only.

### Multiple Imputation with `efa_mi()`

The alternative is to impute the missing values several times, fit each completed dataset,
and pool the results. `EFAtools` does not impute the data itself — use a dedicated tool such
as the [mice](https://CRAN.R-project.org/package=mice) package — but `efa_mi()` takes the
list of completed datasets and does the factor-analytic pooling.

Here we create five imputations with mice (a Bayesian linear-regression model, appropriate
for these continuous items) and collect them into a list.

```{r, eval = mice_ok}
imp <- mice::mice(as.data.frame(d_miss), m = 5, method = "norm",
                  printFlag = FALSE, seed = 123)
dat_list <- lapply(seq_len(imp$m), function(i) mice::complete(imp, i))
```

`efa_mi()` fits the same `efa_fit()` model to each imputed dataset — the extraction,
rotation, and standard-error options are passed through `...` — aligns the solutions to a
common factor space (rotation is only identified up to reflection and permutation, so the
imputations must be matched before averaging), and pools them.

<!-- message = FALSE here only: efa_fit() notes that it is computing correlations from raw
data once per imputation, and five identical notes add nothing. -->

```{r, eval = mice_ok, message = FALSE}
efa_pooled <- efa_mi(dat_list, n_factors = 3, estimator = "ml", rotation = "oblimin")
efa_pooled
```

The pooled loadings recover the three factors. Point estimates are averaged across the
imputations after alignment. The model chi-square, and the AIC/BIC derived from it, are
pooled with the D2 rule — which is why the printout labels the pooled chi-square as a
D2 statistic with its own reference distribution. RMSEA is pooled by the same rule, but
from the discrepancies on the uncorrected N - 1 scale on which it is defined, so it
does not reconcile by hand with the printed chi-square. The incremental CFI and TLI are
instead averaged across the per-imputation fits. Requesting standard errors in the call (for
example `se = "information"` or `se = "np-boot"`) additionally pools them with Rubin's rules,
so the between-imputation variability inflates the pooled standard errors. Because multiple
imputation propagates the extra uncertainty from the missing data, its pooled fit statistics
are not directly comparable with the single FIML fit above; read them together with the
per-imputation fits stored in the returned object.

Which route to prefer is largely practical. FIML is a single, efficient fit and is the
simpler default when the analysis model is the whole story. Multiple imputation is more
flexible when the imputation model should draw on auxiliary variables not in the factor
model, or when the same imputations feed several downstream analyses.

## Ordinal *and* Missing Data

Questionnaire data are usually both: a few response categories *and* some unanswered items.
The two treatments above do not simply combine, because the ordinal machinery needs complete
cases exactly where the FIML route needs continuous ones. We draw the same four-category
items as before, now with 8% of the responses missing completely at random.

```{r}
d_ord_miss <- efa_simulate(N = 300, Lambda = Lambda, Phi = Phi,
                           categories = 4, match = "polychoric",
                           missing = "MCAR", missing_prop = 0.08, seed = 2024)$data

round(mean(is.na(d_ord_miss)), 3)     # overall proportion missing
sum(complete.cases(d_ord_miss))       # respondents who answered every item
```

Eight percent per item is mild, but spread over 18 items it leaves only a fifth of the
sample complete. That number, not the 300 rows supplied, is what the ordinal route has to
work with.

### The Effective Sample of a Polychoric DWLS Fit

Polychoric correlations can be estimated pair by pair, using every respondent who answered
a given pair of items. Their asymptotic covariance cannot: the DWLS weights and the
sandwich standard errors describe *one* set of estimates from *one* set of cases, so
whenever they are requested the correlations and the covariance are both computed on the
listwise-complete rows. `efa_fit()` announces that override rather than applying it
silently.

```{r}
efa_ord_miss <- efa_fit(d_ord_miss, n_factors = 3, cor_method = "poly",
                        estimator = "dwls", rotation = "oblimin")
efa_ord_miss$settings$N        # cases the fit is actually based on
```

Every loading, fit index, standard error, and confidence interval of that fit rests on the
complete cases. `efa_screen()` names both denominators for the same reason — its
multivariate-normality and outlier results state how many complete cases they used out of
how many rows were supplied — so the reduction is visible before a model is fitted.

Two adjustments avoid it, each with a cost:

- Ask for no asymptotic covariance. With `estimator = "ULS"` or `"ML"` and no sandwich
  standard errors, nothing needs the weight matrix, so the polychoric matrix stays
  pairwise-complete and every case contributes to the pairs it answered. In exchange you
  give up the DWLS weighting and the robust standard errors, and the reported `N` is the
  nominal row count while each correlation rests on its own smaller subset — so fit
  statistics and analytic standard errors treat the data as more complete than they are.
- Switch to FIML — but only as a continuous analysis. `cor_method = "fiml"` estimates a
  saturated *normal* covariance and analyses the resulting Pearson-type correlation; there
  is no polychoric asymptotic covariance in it, so it cannot supply DWLS with weights.
  Combining the two is refused rather than silently approximated:

```{r, error = TRUE}
efa_fit(d_ord_miss, n_factors = 3, cor_method = "fiml", estimator = "dwls")
```

### Keeping Both the Ordinal Treatment and the Cases

Multiple imputation is the route that keeps both. Each imputed dataset is complete, so the
polychoric correlations and their asymptotic covariance are estimated on all 300 cases, and
the imputation uncertainty is carried into the pooled results. Since `efa_mi()` forwards its
arguments to `efa_fit()`, this is the ordinal fit above applied to each imputation — here
with predictive mean matching, which resamples observed values and so respects the response
scale.

<!-- message = FALSE here only: the same per-imputation note as above. -->

```{r, eval = mice_ok, message = FALSE}
imp_ord <- mice::mice(as.data.frame(d_ord_miss), m = 5, method = "pmm",
                      printFlag = FALSE, seed = 123)
dat_ord <- lapply(seq_len(imp_ord$m), function(i) mice::complete(imp_ord, i))

efa_ord_pooled <- efa_mi(dat_ord, n_factors = 3, cor_method = "poly",
                         estimator = "dwls", rotation = "oblimin")
efa_ord_pooled
```

The pooled pattern matrix is estimated from all 300 respondents rather than the 62 complete
ones — compare it with `efa_ord_miss$rot_loadings` — and the between-imputation variability
enters any pooled standard errors requested in the call. As always with imputation, the
imputation model has to be defensible: for ordinal items that means a method returning
values on the observed scale (predictive mean matching, or an ordinal model such as
`method = "polr"`), and enough respondents per item to estimate it.

So, for ordinal items with missing values: use `efa_mi()` when the missingness is more than
incidental and the ordinal treatment matters; fit `cor_method = "poly"` with
`estimator = "DWLS"` directly when the complete-case sample is still comfortably large; and
use FIML when the items have enough categories to be treated as continuous in the first
place.

## Where to Next

This vignette covered the ordinal and missing-data extensions of the workflow. For the core
analysis — screening, factor retention, extraction and rotation, and the post-processing
tools — see the [EFAtools](EFAtools.html) vignette, and the individual help pages for the
statistical details and references. Run `browseVignettes("EFAtools")` for the vignettes
installed with the package, or visit the
[package website](https://mdsteiner.github.io/EFAtools/) for the full set of articles.
