---
title: "Multigroup exploratory factor analysis"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 7,
  fig.align = "center"
)
```

When the same set of items is administered to several groups -- age bands, treatment arms,
translated versions of a questionnaire -- a natural question is whether the items measure
the *same* factors in each group. `efa_group()` answers it by fitting an exploratory factor
analysis in every group at a common number of factors, bringing the per-group solutions into
one shared orientation, and then comparing their loadings.

The comparison is not as simple as extracting a solution in each group and reading the
loadings side by side. A factor solution is identified only up to a rotation of its factors,
so two groups can have identical factor structures yet return loading matrices that look
nothing alike -- the factors may be reflected, reordered, or rotated relative to one another.
Before the loadings can be compared they must be brought into a common orientation. That
alignment is what `efa_group()` does for you, on top of the per-group fits, and it is the
reason a dedicated function is worth having.

```{r}
library(EFAtools)
```

## Two groups with a known difference

To see the tools work we need groups whose factor structures we already know, so we can check
that the analysis recovers the difference we built in. We simulate two groups from a
three-factor model with `efa_simulate()`. Both groups share the same simple structure --
three items per factor -- except in how factor 3 is built: in the **control** group item `V7`
is that factor's strongest indicator, whereas in the **treatment** group `V7` nearly drops off
it and `V9` becomes the strongest instead. Factor 3 is therefore defined by a different item in
each group, which should lower its cross-group congruence and light up the per-item difference
flags -- while factors 1 and 2, whose items are untouched, stay highly congruent.

```{r}
# Common simple structure: three items per factor.
Lambda_control <- matrix(0, nrow = 9, ncol = 3,
                         dimnames = list(paste0("V", 1:9), paste0("F", 1:3)))
Lambda_control[1:3, 1] <- c(.70, .65, .60)
Lambda_control[4:6, 2] <- c(.70, .65, .60)
Lambda_control[7:9, 3] <- c(.70, .65, .60)

# The treatment group is identical, except factor 3 is anchored by a different item:
# V7 nearly drops off it and V9 becomes its strongest indicator.
Lambda_treatment <- Lambda_control
Lambda_treatment["V7", 3] <- .20   # was .70
Lambda_treatment["V9", 3] <- .75   # was .60

Lambda_control
Lambda_treatment
```

Each group is drawn with its own seed so the example is fully reproducible.

```{r}
control   <- efa_simulate(N = 300, Lambda = Lambda_control,   seed = 1)$data
treatment <- efa_simulate(N = 300, Lambda = Lambda_treatment, seed = 2)$data

# efa_group() requires the same items, in the same order, in every group.
colnames(control)   <- rownames(Lambda_control)
colnames(treatment) <- rownames(Lambda_treatment)

groups <- list(control = control, treatment = treatment)
```

Groups can be supplied either as a single raw data set together with a `groups` vector (one
value per row), or -- as here -- as a named list of per-group data sets. The list may also
hold per-group correlation matrices instead of raw data, in which case you pass their sample
sizes through `N`.

## Fitting at a common number of factors

`efa_group()` fits each group with `efa_fit()` at the common `n_factors`, forwarding any
estimation and rotation settings unchanged to every group. Here we extract three factors with
a varimax rotation.

```{r}
mg <- efa_group(groups, n_factors = 3, rotation = "varimax")
mg
```

The printed report recaps the groups and the shared model, then summarises the two things we
came for: the factor congruence between the aligned groups, and the size of their loading
differences. The aligned per-group loading matrices are in `$loadings`, sharing one column
order and sign so they can be read against each other directly.

```{r}
lapply(mg$loadings, round, 2)
```

Factors 1 and 2 line up closely across the groups. Factor 3 does not: `V7` and `V9` have
swapped prominence, exactly as we built in.

## Alignment: consensus versus reference

`efa_group()` chooses one of two alignment strategies automatically, and the report tells you
which one it used.

* **Consensus** (the default for orthogonal and unrotated solutions) builds a single symmetric
  target across *all* groups by generalized Procrustes analysis and rotates every group to it.
  No group is privileged -- this is what the fit above used.
* **Reference** keeps one group's loadings fixed and rotates every other group onto them. This
  path is used whenever you name a `reference_group`, and it is used *automatically* for
  oblique rotations, because the symmetric consensus target is not defined for oblique
  transformations with more than one factor.

When an oblique rotation triggers the reference path without an explicit choice, `efa_group()`
reports which group it used and leaves the requested rotation untouched:

```{r, message = TRUE}
mg_obl <- efa_group(groups, n_factors = 3, rotation = "oblimin")
mg_obl$settings$alignment
```

Supplying `reference_group` forces the reference path and chooses the anchor yourself -- by
name or by position:

```{r}
mg_ref <- efa_group(groups, n_factors = 3, rotation = "varimax",
                    reference_group = "control")
mg_ref$settings$alignment
```

## Factor congruence with bootstrap intervals

Cross-group factor similarity is quantified by Tucker's congruence coefficient between the
aligned loadings -- one value per matched factor, per group pair. A congruence near 1 means
the factor is defined by the same items, in the same relative proportions, in both groups.

With raw data you can attach a percentile bootstrap confidence interval to each congruence by
setting `b_boot`. The bootstrap resamples cases within each group, refits, re-aligns every
replicate to the frozen target, and recomputes the congruences; a `seed` makes it reproducible
and independent of any parallel backend.

```{r}
mg_ci <- efa_group(groups, n_factors = 3, rotation = "varimax",
                   b_boot = 500, seed = 2026)

mg_ci$congruence$matched["control", "treatment", ]
mg_ci$congruence$matched_ci$lower["control", "treatment", ]
mg_ci$congruence$matched_ci$upper["control", "treatment", ]
```

The `plot()` method draws the matched congruences against the Lorenzo-Seva and ten Berge
(2006) similarity bands -- `.95` and above reads as "equal", `.85` to `.95` as "fair" -- so a
factor's cross-group similarity can be judged at a glance. With a bootstrap the points carry
their confidence intervals.

```{r, fig.height = 4}
plot(mg_ci, type = "congruence")
```

Factors 1 and 2 sit at the top of the scale; factor 3, whose indicators were rearranged, drops
into or below the "fair" band.

## Loading differences: salient versus significant

Congruence summarises a whole factor. To see *which items* drive a difference, `efa_group()`
also compares the aligned loadings cell by cell. `$diffs` gives a per-pair magnitude summary,
and `$flags` gives the signed difference for every item on every factor.

An item is **flagged** when the groups' aligned loadings differ by at least `delta` (default
`0.1`) in absolute value. This is a descriptive *salience* threshold, not a significance test.
When a bootstrap was run, each difference additionally carries a confidence interval, and
`ci_excludes_0` reports whether that interval clears zero -- the closest thing to a
significance statement. The two need not agree: a difference can be large enough to flag yet
have an interval spanning zero (salient but uncertain), or small yet reliably non-zero.

```{r}
mg_ci$diffs
```

The factor-3 cells for `V7` and `V9` stand out. Restricting `$flags` to the flagged cells
shows where the two notions of "different" line up:

```{r}
flagged <- mg_ci$flags[mg_ci$flags$flagged, ]
flagged[, c("indicator", "factor", "diff", "ci_excludes_0")]
```

The `"differences"` plot lays the signed differences out as a heatmap, one panel per group
pair, outlining the cells that reach `delta`.

```{r, fig.height = 4.5}
plot(mg_ci, type = "differences")
```

## An approximate-invariance verdict

For a compact summary, `invariance = TRUE` grades each factor's congruence against the
Lorenzo-Seva and ten Berge (2006) bands and returns a per-factor verdict: `"equal"` for a
congruence of at least `.95`, `"fair"` for `.85` to `.95`, and `"incongruent"` below that.

```{r}
mg_inv <- efa_group(groups, n_factors = 3, rotation = "varimax",
                    b_boot = 500, seed = 2026, invariance = TRUE)
mg_inv$invariance
```

When a bootstrap is available the verdict is read conservatively off the *lower* bound of the
congruence interval, so a factor is graded "equal" only if even the lower bound clears the
band. That is why factors 1 and 2 above -- comfortably above `.95` as point estimates -- are
graded only "fair": their interval lower bounds slip below `.95`. This is deliberately
cautious, guarding against declaring invariance on the strength of a point estimate that a
modest resample could pull below the threshold. Without a bootstrap the verdict falls back to
the point estimate, and the `phi_lower` column is left `NA` to signal the non-conservative
reading.

The verdict is a graded heuristic built on the congruence bands, not a formal test of
measurement invariance. It is a fast, exploratory read on how comparable the factors are across
groups -- a confirmatory multigroup model remains the tool for a formal invariance test.

## Where to next

`efa_group()` carries several further controls: `delta` to tune the salience threshold, `ci`
to set the interval level, and any `efa_fit()` argument (`estimator`, `cor_method`, an
`estimate_control()` or `rotate_control()`) to configure the per-group estimation. See
`?efa_group` for the full argument list and the structure of the returned object, and the
[EFAtools](EFAtools.html) vignette for the single-group workflow the per-group fits build on.

## References

Gower, J. C. (1975). Generalized Procrustes analysis. *Psychometrika*, 40, 33-51.

Lorenzo-Seva, U., & ten Berge, J. M. F. (2006). Tucker's congruence coefficient as a
meaningful index of factor similarity. *Methodology*, 2, 57-64.
